From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-08 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Comments and modifications from meeting

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, the density itself is coordinate-dependent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

@@

Our argument has three main steps, summarized in Figure~\ref{diagram}. The left column lists three basic physical conditions: infinitesimal reducibility, determinism/reversibility, and motion equivalence. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal mathematical framework depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby gain a straightforward interpretation of the physical significance of these formalisms.
%Of the various notions of fundamentality, there is at least the following non-mysterious definition:

@@

If a system is infinitesimally reducible, then we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that to define this invariant distribution, we require a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space. Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
% [To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises ONLY when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

Of course, understanding the relationship between Hamiltonian and Lagrangian mechanics partly depends on how one defines ``classical systems." Different definitions can result in different interpretive relationships. On our account, Newtonian systems are the widest domain of classical systems, including any system that is infinitesimally reducible. Hamiltonian mechanics picks out the subset of Newtonian systems that follow deterministic and reversible evolution. Our treatment thereby sets aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative, i.e. conserving energy. Consequently, our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although the Lagrangian formulation applies to dissipative systems \parencites[]{Smith,Curiel}, this is not what we will mean by ``Lagrangian systems."\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian. Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by ``fundamental."

@@

Owing to our epistemic inabilities to access these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant distributions (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus structure. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. Precisely this scenario occurs in the context of Lagrangian mechanics, which requires the additional condition of motion equivalence to hold. Whereas North views Lagrangian mechanics as containing ``excess, superfluous structure" \parencites*[75]{North}, we take this additional condition as unnecessary but still physically significant. Thus, while it is illuminating to identify the minimum mathematical structures necessary for a formulation, additional structure is not automatically rendered physically insignificant. %It corresponds to the physical assumption of motion equivalence, which applies to many classical systems. 
%This characterizes the relationship between Hamiltonian and Lagrangian formalizations of a classical system. 
% classical mechanics also deploys physically significant representations that are not intrinsic to mechanical systems [to do: mention some examples, e.g. in a footnote. The Lagrangian? Cartesian coordinates?].

@@

First, we will show that a classical phase space arises if and only if a system satisfies the following physical condition:

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any point $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point in $\mathcal{S}$ describes a possible state of a classical particle. We assume that this state space $\mathcal{S}$ is a manifold.\footnote{Ultimately, we derive an if-then claim: if the state space of classical particles is a manifold, then it must be a symplectic manifold.} Then for each state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state. Formally, if $\mathcal{S}$ is a manifold, $\int_U \rho_c d\mathcal{S}$ gives us the fraction of the system whose parts are in the region $U$ of the state space $\mathcal{S}$. 
%[is region U in the state space C or S. I assume in S.].

We define the \textit{state variables} as a set of quantities $\xi^a$ that fully identify a state. Formally, $\xi^a: \mathcal{S} \to \mathbb{R}^n$; this function yields a real-valued vector for any given particle state, possessing $n$-many degrees of freedom.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates,' but we will reserve that terminology for the canonical position coordinates, which are the state variables that define physical units.} We will define a \textit{coordinate} $q \in \xi^a$ as a particular state variable that defines a unit.  For instance, $q$ might be position, defining units of meters. In general, the coordinates are the subset of the state variables that define the unit system, needed to measure properties of the system.

@@

On physical grounds, the distribution $\rho_c$ must transform as both a density and a scalar function. As a function defined over the state variables, the distribution transforms as a density, i.e. $\rho_c$ changes as a function of the Jacobian determinant that relates new state variables to old ones. As a function defined over the points in phase space, the distribution transforms as a scalar, i.e. $\rho_c$ is invariant under all state variable transformations. Hence, the Jacobian determinant must in fact equal one. This means that $\rho_c$ is a function only of the particle state $s$ and does not depend on the choice of coordinates.\footnote{Formally, we can express this argument as a chain of equalities: $\rho_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_c(\hat\xi^b)$, where $\rho_c(\xi^a)$ and $\rho_c(\hat{\xi}^b)$ represent the same distribution as expressed over different state variables and $s(\xi^a)$ and $s(\hat{\xi}^b)$ represent the same particle state as expressed by the different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through the particle state.} We will now show that in order for the distribution to satisfy these two constraints, the classical phase space must be at least a symplectic manifold. In other words, having symplectic manifold structure is a necessary condition for the distribution over particle states to transform as a density and scalar function.
%This is because the distribution is a function only of the states, rather than of the particular choice of coordinates. 
 % Classical phase space (i.e. a symplectic manifold) is the only space that satisfies these constraints on the distribution. 
%(in manifold terminology, these are known as coordinates)

We begin with the simplest case: suppose that a single coordinate $q $ suffices to define the unit system for the particle state space. Then, the state variables are given by a set $\xi^a = \{ q, k^i \}$, where \textit{prima facie} the index $i$ can range over more than one number. We will show that, in fact, there can only be a single $k^i$ variable in this case, conjugate to the single $q$ coordinate. Assuming that the particle density $\rho_c$ is a scalar, it follows that we can arbitrarily change the unit defined by $q$ without changing the density $\rho_c$. 
% We further assume that $q$ fully defines the units for all state variables. [Is there a way to physically motivate this assumption, beyond this being a constitutive role of the position-like state variables?]

@@

The conclusion is that, under the infinitesimal reducibility assumption, a state $c$ of the system is a coordinate-invariant distribution over the state space $\mathcal{S}$ of its infinitesimal parts. If the state space of the particles is a manifold, then it must be a symplectic manifold: symplectic manifold structure is necessary for the distribution to be coordinate-invariant. 
% Could connect this with the earlier discussion of North's methodology of seeking coordinate-invariants, from section~\ref{fundamentality}.

@@

If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space. Note that each space variable $x^i$ is a coordinate, i.e. a state variable that defines a unit. In fact, the trajectories can be fully described by these units and only these units. For, to specify a trajectory, we need epistemic access to a unit system, where the number of independent units is determined by the dimensionality of physical space. It follows that $x^i = q^i$, each $q^i$ is paired with a conjugate $p_i$, and each state $\{q^i, p_i\}$ corresponds to one and only one trajectory in physical space. At each point $x^i$, then, there must be infinitely many trajectories, one for each possible combination of $\{p_i\}$. Since these trajectories are differentiable in $x^i$, we can define a velocity $v^i = d_t x^i$. If the equations of motions were such that $v^i=v^i(q^i)$, then motion equivalence would fail as the full trajectory would be determined only by the $q^i$. Hence, we must have $v^i=v^i(q^j, p_k)$, so that knowledge of the state in phase space suffices for knowledge of the trajectory. If motion equivalence holds, then this relationship must be invertible, meaning that knowledge of the trajectory suffices for knowledge of the phase space state. For systems that satisfy the motion equivalence condition, then, the following relationships hold at any given time:
\begin{equation}\label{weak_equivalence}
\begin{aligned}
x^i &= q^i \\
v^i &= d_t x^i = v^i(q^j, p_k)
\end{aligned}
\end{equation}

@@

Next, we show that this Lagrangian describes massive particles under scalar potentials. If we look at the whole distribution (so that $q$ is no longer held fixed), we have $\rho(q^i, p_j) = |J| \rho(x^i, v^j) = \left|\frac{\partial v^i}{\partial p_j}\right| \rho(x^i, v^j)$ since
\begin{equation}
\begin{aligned}
|J| &= \begin{vmatrix}
\frac{\partial x^i}{\partial q^j} & \frac{\partial x^i}{\partial p_j} \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix}
= \begin{vmatrix}
\delta^i_j & 0 \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix} \\
&= \left|\delta^i_j\right| \left|\frac{\partial v^i}{\partial p_j}\right| 
= \left|\frac{\partial v^i}{\partial p_j}\right|.
\end{aligned}
\end{equation}
Note that while the value given by $\rho(q^i, p_j)$ is coordinate independent, the value given by $\rho(x^i, v^j)$ depends on the choice of coordinate through $\left|\frac{\partial v^i}{\partial p_j}\right|$. If $x^i$ truly sets the unit system by itself, then $\left|\frac{\partial v^i}{\partial p_j}\right|$ must be a function of position only. Similar considerations will also hold for marginal distributions (i.e. distributions on a subset of the coordinates), which means all components of $\frac{\partial v^i}{\partial p_j}$ must be a function of position only. We set:\footnote{In Section~\ref{dependence}, we will see that $g^{ij}$ corresponds to the Riemannian metric, since $\partial v^i/\partial p_j$ it is a function only of position.}
\begin{equation}
\begin{aligned}
\frac{\partial v^i}{\partial p_j} = \frac{1}{m} g^{ij} \\
\frac{\partial p_j}{\partial v^i} = m g_{ji}
\end{aligned}
\end{equation}
where $m$ is the unit conversion constant between velocity and conjugate momentum, while $g_{ij}$ represents the linear dependency of the velocity on the momentum.

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics.
% Canonical coordinates are physically significant because they allow us to define an invariant phase space density.
% % This latter claim might seem puzzling, since insofar as we need to use canonical coordinates to define the density, this seems to speak in favor of the canonical coordinates being more fundamental.

@@

Hence, the density is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not generically coordinate invariant, since the density changes with the Jacobian. For example, changing from polar to Cartesian coordinates, we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, the value of $|J|$ can change from point to point. Consequently, the density is not guaranteed to be directly comparable at different points: point-wise comparison need not hold in an arbitrary coordinate system. If we have a uniform, constant density in $x$ and $y$, we will have an $r$-dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$. This makes the value of the density, in general, not directly comparable at different points. Hence, to determine whether there is more stuff at one location in state space than another, we generically need to know which coordinate system we are in, and how the density depends on these coordinates (i.e. we can't just compare the values of $\rho$ in these coordinates and see which is larger). 
%The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.
%when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). [see 57.3 minutes of 12-21 recording for discussion]

>>>>

Subsequent sections describe additional advantages of Hamiltonian mechanics, along with further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of matter in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, the density itself is coordinate-dependent. [XXX: one cannot make density truly coordinate independent] We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

@@

Our argument has three main steps, summarized in Figure~\ref{diagram}. The left column lists three basic physical conditions: infinitesimal reducibility, determinism/reversibility, and motion equivalence. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal/mathematical structure depicted in the right column, including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby gain a straightforward interpretation of the physical significance of these formalisms.
%Of the various notions of fundamentality, there is at least the following non-mysterious definition:

@@

If a system is infinitesimally reducible, then we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that defining this invariant distribution requires a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

@@

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space.\footnote{XXX: Since photons go at constant speed, one can only recover the direction of momentum from the trajectory, but not the magnitude. Or equivalently, two photons may still follow the same trajectory if they only differ in the magnitude of momentum.} Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
% [To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises ONLY when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

Of course, understanding the relationship between Hamiltonian and Lagrangian mechanics partly depends on how one defines ``classical systems." Different definitions can result in different interpretive relationships. On our account, Newtonian systems [XXX: Nope!] are the widest domain of classical systems, including any system that is infinitesimally reducible. Hamiltonian mechanics picks out the subset of Newtonian systems that follow deterministic and reversible evolution. Our treatment thereby sets aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \textcites*[3-4]{Baez}{Earman}{Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative, i.e. conserving energy. Consequently, our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although the Lagrangian formulation applies to dissipative systems \parencites[]{Smith,Curiel}, this is not what we will mean by ``Lagrangian systems."\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian. Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by ``fundamental."

% https://phys.libretexts.org/Bookshelves/Classical_Mechanics/Book%3A_Variational_Principles_in_Classical_Mechanics_(Cline)/10%3A_Nonconservative_Systems/10.04%3A_Rayleighs_Dissipation_Function

@@

Owing to our epistemic inabilities to access these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant distributions (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus structure. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. Precisely this scenario occurs in the context of Lagrangian mechanics, which requires the additional condition of motion equivalence to hold. Whereas North views Lagrangian mechanics as containing ``excess, superfluous structure" \parencites*[75]{North}, we take this additional condition as optional but still physically significant. Thus, while it is illuminating to identify the minimum mathematical structures necessary for a formulation, additional structure is not automatically rendered physically insignificant. %It corresponds to the physical assumption of motion equivalence, which applies to many classical systems. 
%This characterizes the relationship between Hamiltonian and Lagrangian formalizations of a classical system. 
% classical mechanics also deploys physically significant representations that are not intrinsic to mechanical systems [to do: mention some examples, e.g. in a footnote. The Lagrangian? Cartesian coordinates?].

@@

First, we will show that a classical phase space arises if a system satisfies the following physical condition:

@@

Let $\mathcal{C}$ be the state space for the whole system, so that any point $c \in \mathcal{C}$ represents a particular state of the whole system. Given infinitesimal reducibility, a classical system consists of a collection of classical particles. Let $\mathcal{S}$ be the state space for the particles, so that each point in $\mathcal{S}$ describes a possible state of a classical particle. We assume that this state space $\mathcal{S}$ is a manifold.\footnote{Ultimately, we derive an if-then claim: if the state space of classical particles is a manifold, then it must be a symplectic manifold.} [XXX: if time is continuous, trajectories in state space must be continuous, local structure must be compatible] Then for each state $c \in \mathcal{C}$ there exists a unique distribution $\rho_c : \mathcal{S} \to \mathbb{R} $ that describes the state of $c$'s parts. That is, the state of the whole system is a distribution over its infinitesimal parts. This distribution describes the proportion of classical particles in any particular state. Formally, if $\mathcal{S}$ is a manifold, $\int_U \rho_c d\mathcal{S}$ gives us the fraction of the system whose parts are in the region $U$ of the state space $\mathcal{S}$. 
%[is region U in the state space C or S. I assume in S.].

We define \textit{state variables} as a set of quantities $\xi^a$ that fully identify a state. Formally, $\xi^a: \mathcal{S} \to \mathbb{R}^n$; this function yields a real-valued vector for any given particle state, possessing $n$-many degrees of freedom.\footnote{In the language of manifolds, these $\xi^a$ are called `coordinates,' but we will reserve that terminology for the canonical position coordinates, which are the state variables that define physical units. [XXX: coordinates -> unit variables]} We will define a \textit{coordinate} $q \in \xi^a$ as a particular state variable that defines a unit.  For instance, $q$ might be position, defining units of meters. In general, the coordinates are the subset of the state variables that define the unit system, needed to measure properties of the system.

@@

On physical grounds, [XXX: I'd say it's more mathematical grounds] the distribution $\rho_c$ must transform as both a density and a scalar function. As a function defined over the state variables [XXX: within the integral $\int_U \rho_c(\xi^a) d\xi^a$], the distribution transforms as a density, i.e. $\rho_c$ changes as a function of the Jacobian determinant that relates new state variables to old ones. As a function defined over the points in phase space [XXX:$\rho_c : \mathcal{S} \to \mathbb{R} $], the distribution transforms as a scalar, i.e. $\rho_c$ is invariant under all state variable transformations. Hence, the Jacobian determinant must in fact equal one. This means that $\rho_c$ is a function only of the particle state $s$ and does not depend on the choice of coordinates.\footnote{Formally, we can express this argument as a chain of equalities: $\rho_c(\hat{\xi}^b)=\rho_c(s(\hat{\xi}^b))=\rho_c(s(\xi^a))=\rho_c(\xi^a) = \left|\frac{\partial \hat{\xi}^b}{\partial \xi^a} \right| \rho_c(\hat\xi^b)$, where $\rho_c(\xi^a)$ and $\rho_c(\hat{\xi}^b)$ represent the same distribution as expressed over different state variables and $s(\xi^a)$ and $s(\hat{\xi}^b)$ represent the same particle state as expressed by the different variables. Note that $\rho_c(s(\hat{\xi}^b))$ shows how the distribution depends on the state variables only through the particle state.} We will now show that in order for the distribution to satisfy these two constraints, the classical phase space must be at least a symplectic manifold. In other words, having symplectic manifold structure is a necessary condition for the distribution over particle states to transform as a density and scalar function.
%This is because the distribution is a function only of the states, rather than of the particular choice of coordinates. 
 % Classical phase space (i.e. a symplectic manifold) is the only space that satisfies these constraints on the distribution. 
%(in manifold terminology, these are known as coordinates)

We begin with the simplest case: suppose that a single coordinate $q $ suffices to define the unit system for the particle state space. Then, the state variables are given by a set $\xi^a = \{ q, k^i \}$, where \textit{prima facie} the index $i$ can range over more than one number. [XXX: or less] We will show that, in fact, there can only be a single $k^i$ variable in this case, conjugate to the single $q$ coordinate. Assuming that the particle density $\rho_c$ is a scalar, it follows that we can arbitrarily change the unit defined by $q$ without changing the density $\rho_c$. 
% We further assume that $q$ fully defines the units for all state variables. [Is there a way to physically motivate this assumption, beyond this being a constitutive role of the position-like state variables?]

@@

The conclusion is that, under the infinitesimal reducibility assumption, a state $c$ of the system is a coordinate-invariant distribution over the state space $\mathcal{S}$ of its infinitesimal parts. If the state space of the particles is a manifold, then it must be a symplectic manifold: symplectic manifold structure is necessary for the distribution to be coordinate-invariant. Additionally, any symplectic manifold structure will allow the definition of coordinate--invariant distributions: any scalar function can be in fact integrated using the symplectic form in a coordinate-invariant way. Therefore the symplectic structure is exactly the structure needed for their definition. [XXX:]
% Could connect this with the earlier discussion of North's methodology of seeking coordinate-invariants, from section~\ref{fundamentality}.

@@

If we focus on the parts of the system, this means that for every evolution in phase space there should be one and only one trajectory in physical space. Note that each space variable $x^i$ is a coordinate, i.e. a state variable that defines a unit. In fact, the trajectories can be fully described by these units and only these units. For, to specify a trajectory, we need epistemic access to a unit system, where the number of independent units is determined by the dimensionality of physical space. It follows that $x^i = q^i$, each $q^i$ is paired with a conjugate $p_i$, and each state $\{q^i, p_i\}$ corresponds to one and only one trajectory in physical space. At each point $x^i$, then, there must be infinitely many trajectories, one for each possible combination of $\{p_i\}$. Since these trajectories are differentiable in $x^i$, we can define a velocity $v^i = d_t x^i$. If the equations of motions were such that $v^i=v^i(q^i)$, then motion equivalence would fail: since both $x^i$ and $v^i$ are a function of only $q^i$, the relationship would not be invertible because of dimensional considerations. [XXX] Hence, we must have $v^i=v^i(q^j, p_k)$, so that knowledge of the state in phase space suffices for knowledge of the trajectory. If motion equivalence holds, then this relationship must be invertible, meaning that knowledge of the trajectory suffices for knowledge of the phase space state. For systems that satisfy the motion equivalence condition, then, the following relationships hold at any given time:
\begin{equation}\label{weak_equivalence}
\begin{aligned}
x^i &= q^i \\
v^i &= d_t x^i = v^i(q^j, p_k)
\end{aligned}
\end{equation}

@@

Note that weak equivalence is not enough. The classical state, as we defined it, is really the composite, the distribution, and we must be able to express that distribution in terms of the position and velocity. [XXX] To have full equivalence, we must be able to express the density as well. Therefore we must have $\rho(q^i, p_j) = |J| \rho(x^i, v^j) = \left|\frac{\partial v^i}{\partial p_j}\right| \rho(x^i, v^j)$ since
\begin{equation}
\begin{aligned}
|J| &= \begin{vmatrix}
\frac{\partial x^i}{\partial q^j} & \frac{\partial x^i}{\partial p_j} \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix}
= \begin{vmatrix}
\delta^i_j & 0 \\
\frac{\partial v^i}{\partial q^j} & \frac{\partial v^i}{\partial p_j}
\end{vmatrix} \\
&= \left|\delta^i_j\right| \left|\frac{\partial v^i}{\partial p_j}\right| 
= \left|\frac{\partial v^i}{\partial p_j}\right|.
\end{aligned}
\end{equation}
Note that while the value given by $\rho(q^i, p_j)$ is invariant under unit, the value given by $\rho(x^i, v^j)$ depends on the choice of units through $\left|\frac{\partial v^i}{\partial p_j}\right|$. If $x^i$ truly sets the unit system by itself, then $\left|\frac{\partial v^i}{\partial p_j}\right|$ must be a function of position only, since the position variables, by themselves, define the units. Similar considerations will also hold for marginal distributions (i.e. distributions on a subset of the coordinates), which means all components of $\frac{\partial v^i}{\partial p_j}$ must be a function of position only. We set:\footnote{In Section~\ref{dependence}, we will see that $g^{ij}$ corresponds to the Riemannian metric, since $\partial v^i/\partial p_j$ it is a function only of position.}
\begin{equation}
\begin{aligned}
\frac{\partial v^i}{\partial p_j} = \frac{1}{m} g^{ij} \\
\frac{\partial p_j}{\partial v^i} = m g_{ji}
\end{aligned}
\end{equation}
where $m$ is the unit conversion constant between velocity and conjugate momentum, while $g_{ij}$ represents the linear dependency of the velocity on the momentum.

@@

Section~\ref{dependence} shows why point-wise comparison does not hold in general: the density is a coordinate-dependent quantity, and thus it is not generically coordinate-invariant. Consequently, within the Lagrangian formalism, point-wise comparison holds only when we have made a special choice of coordinates. This gives a further reason to privilege Hamiltonian mechanics. Section~\ref{fundamental} argues for the fundamental physical significance of the phase space density. Since the density is coordinate-dependent, this shows that coordinate-independence is not a necessary condition for fundamental physical significance. Our position thereby undercuts metaphysically-driven demands to represent physical theories in exclusively ``intrinisic" terms.\footnote{For arguments in favor of representing physical systems and explanations in exclusively intrinsic terms, see \textcites[]{Field,North,Chen2018}.} Section~\ref{gauge_momentum} shows why the conjugate momentum $p$ is itself a gauge-dependent quantity. This bolsters our claim that the density is more fundamental than state variable pairs. It also shows that---\textit{pace} conventional wisdom---position and conjugate momentum are \textit{not} on equal physical footing within Hamiltonian mechanics.\footnote{This illustrates how one can be mislead by the syntactical form of the mathematics to draw an interpretive conclusion that is not borne out by the physics.[XXX]}
% Canonical coordinates are physically significant because they allow us to define an invariant phase space density.
% % This latter claim might seem puzzling, since insofar as we need to use canonical coordinates to define the density, this seems to speak in favor of the canonical coordinates being more fundamental.

@@

Hence, the density is coordinate dependent because its definition requires us to make a choice of coordinates. Worse than that, it is not generically coordinate invariant, since the density changes with the Jacobian. For example, changing from polar to Cartesian coordinates, we have $\rho(x,y) = |J| \rho(r, \theta) = r \rho(r, \theta)$. Since the Jacobian is a function of the point, the value of $|J|$ can change from point to point. Consequently, the density is not guaranteed to be directly comparable at different points: point-wise comparison need not hold in an arbitrary coordinate system. If we have a uniform, constant density in $x$ and $y$, we will have an $r$-dependence in $r$ and $\theta$. This dependence is introduced precisely because the area identified by $dr$ and $d\theta$ is $dr \,r d\theta$, which is a function of $r$. This makes the value of the density, in general, not directly comparable at different points. [XXX: add remark about size of binning with low statistics?] Hence, to determine whether there is more stuff at one location in state space than another, we generically need to know which coordinate system we are in, and how the density depends on these coordinates (i.e. we can't just compare the values of $\rho$ in these coordinates and see which is larger). 
%The Hamiltonian formalism circumvents this further epistemic requirement by defining the density using quantities that vary covariantly. Hence, we can perform point-wise comparisons in the Hamiltonian formalism without knowing what coordinate system we are in.
%when we define the density originally, it is coordinate-free (rho as a function of the state space of the particles, S). But when we go to express it numerically, then it becomes coordinate-dependent. (I.e. when we express it over a manifold, then it needs to be a density for integration, and hence it becomes coordinate-dependent). [see 57.3 minutes of 12-21 recording for discussion]
